African Wild Dog Dispersal and Connectivity under Climate Change - Lessons Learned from Seasonal Flood Extremes

### 2023-03-14 Chapter_8/04_Manuscript/Manuscript_2023-01-19.tex
various updates

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1,2} \orcid{0000-0001-7477-2642} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940}
}

@@

\textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
Private Bag 13, Maun, Botswana.

@@

\textbf{Running Title:} African Wild Dog Dispersal in a Changing Climate:
Lessons Learned from Seasonal Flood Extremes

@@

\textbf{Keywords:} movement, connectivity, climate change, seasonality,
dispersal, conservation, individual-based simulations, step-selection

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Climate change is expected to profoundly impact the life history of wild-living
animal populations. While the impacts of climate change on the demographics of
local subpopulations have been studied repeatedly, little is known about the
consequences for dispersal and connectivity.

We capitalize on a ``natural experimental setup'', the flood-pulse driven change
in surface-water across the Okavango Delta in northern Botswana, to investigate
the impact of changing environmental conditions on dispersal patterns and
connectivity of the endangered African wild dog (\textit{Lycaon pictus}). For
this, we simulate dispersal trajectories across the Okavango Delta under two
extreme scenarios that serve to represent environmental conditions akin to those
expected under continued climate change; one assuming a maximum flood, one
assuming a minimum flood.

During maximum flood, the Okavango Delta poses an important dispersal barrier
that reduces dispersal prospects in increases dispersal durations between
distinct areas. Across the entire study area, we observe
$\input{99_NumberReachingOthersPercentage.tex}$\% lower dispersal success and
$\input{99_StepsToReachingOthersPercentage.tex}$\% longer dispersal durations
during maximum flood. Most notably, dispersal into the central habitats of the
Okavango Delta is reduced by $\input{99_NumberReachingArea6Percentage.tex}$\%
with an accompanied increase in dispersal durations of
$\input{99_StepsToReachingArea6Percentage.tex}$\%. Depending on the flood,
dispersal corridors traversed different areas and dispersers moved into
proximity of different human-dominated areas.

Whilst the exact impacts of climate change on the flooding regime of the
Okavango Delta remain unknown, our results suggest that connectivity will vastly
differ depending on future flood conditions. Acknowledging such differences will
be key to design effective conservation strategies, especially in light of
ongoing climate change. Since we highlight critical dispersal corridors and
human-wildlife conflict zones for two distinct future scenarios, our results
will facilitate the evidence-based conservation of the endangered African wild
dog.

@@

\subsection{What We Did}
Here, we utilize a previously parameterized and validated dispersal model as
IBMM to simulate dispersal and assess dispersal success and connectivity
patterns for African wild dogs under two extreme scenarios: one assuming maximum
flooding of the Okavango delta and one assuming minimum flooding of the Okavango
delta. The IBMM was trained using GPS data collected during dispersal and
frequently updated environmental data, thus providing a high degree of realism
\citep{Hofmann.2021}. Given that dispersers avoid crossing through water (albeit
we do occasionally observe it in the field), we anticipated that dispersal
prospects and connectivity during maximum flood would be low. Moreover,  when
the flood extent of the OD is at its maximum, the water extends almost into the
densely populated village of Maun. Since both the flood and humans are avoided
by dispersing wild dogs, we anticipated that a fully flooded Delta would result
in a total halt of movement between the Western and Eastern side of the OD.
Information on habitat selection or connectivity is also suitable for predicting
areas with an elevated potential for human wildlife conflict
\citep{Buchholtz.2020}. Hence, also quantified areas with high potential for
human wildlife conflict.

@@

\subsection{Spatial Habitat Layers}
We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each with a resolution of 250 m. The
set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a human influence layer depicting anthropogenic
influences through villages, roads, and agriculture. A detailed description of
the different habitat layers is provided in \citep{Hofmann.2021}. Importantly,
the water-cover and derived distance-to water layers were generated using MODIS
Terra MCD43A4 satellite imagery that was classified into binary water-cover maps
using a ``floodmapping'' algorithm developed by \citep{Wolski.2017}. This
allowed us to generate almost weekly updated ``floodmaps'', thus providing
detailed information about the flood-extent at any given point in time. In
total, we generated 700 floodmaps between the years 2000 and 2019, based on
which we generated a minimal and maximum flood scenario. To create the minimum
(maximum) flood scenario, we averaged the 100 floodmaps with the smallest
(highest) flood extent and generated a binary layer by masking all pixels that
were inundated in less than 50\% of the maps. The resulting maps are presented
in \Cref{FloodExtent}. Ultimately, we combined the habitat layers into two
stacks, one representing the minimum flood scenario, one representing the
maximum flood scenario. To mitigate edge effects during the dispersal
simulation, we followed \cite{Koen.2010} and expanded the spatial extent of the
stacked layers by 20\% and randomized habitat values within the so created
buffer zone (red rectangle in \Cref{StudyArea}a).

@@

\subsection{Dispersal Simulation}
We used a previously parameterized and validated dispersal model to simulate
dispersal of AWDs. The dispersal model was trained using GPS data of 16 wild dog
coalitions dispersing across northern Botswana \citep{Hofmann.2023} which was
fed into an integrated step-selection function (iSSF, \citealp{Avgar.2016}). In
iSSFs, consecutive GPS locations are converted into steps (the straight-line
traveled between two GPS recordings \citep{Turchin.1998a}) and compared to a set
of \textit{random} steps in a conditional logistic regression framework
\citep{Fortin.2005, Thurfjell.2014, Muff.2020, Fieberg.2021}. Because iSSFs
capitalize on the autocorrelated nature of the collected data, they provide
better estimates of connectivity than traditional resource selection approaches
\citep{Zeller.2016}. The model presented in \cite{Hofmann.2023} comprised of a
movement kernel, describing how dispersers move across the landscape in the
absence of habitat selection, a habitat kernel, indicating preferred or avoided
habitat features, and interactions among the two, i.e. how movement behavior
changes depending on habitat conditions. According to this model, the main
characteristics of AWD dispersal movements are avoidance of water, avoidance of
areas influenced by humans, and a preference for directional and fast movements.
The model parameters are provided in Appendix SX and explained in
\citealp{Hofmann.2023}.

Originating from each of the six source areas, we simulated 2'000 individuals
dispersing for a total of 2'000 steps. 1'000 individuals were simulated assuming
a minimum flood, the remaining 1'000 assuming a maximum flood. This resulted in
the simulation of a total of 12'000 individuals. The simulation procedure was
based on the algorithm described in \cite{Hofmann.2023} and works as follows. A
random location within the source area is defined as starting point. Originating
from the starting point, a set of 25 random steps is generated by sampling step
lengths from a gamma distribution fitted to observed steps (shape = 0.37, scale
= 6'316) and turning angles from a uniform distribution ($-\pi, +\pi$). Along
each random step the underlying spatial covariates are extracted, and relevant
movement metrics are computed. \(\beta-\)estimates from the fitted model are
used to predict the probability of each step for being chosen, given the steps
associated covariates. Among the 25 proposed steps, one is chosen at random
based on assigned probabilities. The location of the animal is updated, and the
procedure is repeated until the desired number of steps is realized. Here, we
simulated each individual for 2'000 steps, corresponding to a dispersal duration
of 400 days and the longest dispersal duration recorded in this study area
\citep{Cozzi.2020, Hofmann.2021}. The simulated trajectories can be understood
as correlated random walks.

@@

\section{Results}
Figures depicting the derived connectivity and human-wildlife conflict maps are
provided in \Cref{Metrics}. Difference maps to visualize the differences between
minimum and maximum flood are given in \Cref{Difference}. For brevity, we will
here focus on system-wide connectivity patterns and only selectively point to
regional results. Local connectivity maps derived for each source-area
separately are presented in the Appendix. As the heatmaps in \Cref{Metrics}a
reveal, the OD acts as major dispersal barrier during maximum flood yet reveals
vital dispersal habitat during minimum flood. Differences between maximum and
minimum flood (\Cref{Difference}a) are particularly pronounced for the region
between source areas 1 and 2, where few dispersers occur during times of maximum
flood. In fact, because the floodwaters of the OD reach almost into Maun, the OD
creates a line of separation between its eastern and western sections. The
separation is further amplified as the city of Maun is avoided by dispersers in
both scenarios. Similar patterns are observed on the betweenness maps
(\Cref{Metrics}b), where several pinch-points and bottlenecks linking source
area 6 to the surrounding source areas exist during minimum flood. During
maximum flood, however, these links vanish and instead a single corridor at the
south-eastern tip of the OD emerges (\Cref{Difference}b). Despite its apparent
importance in linking the eastern and western sections of the delta, it is
evident from (\Cref{Metrics}a) that this corridor is only rarely used,
especially during the maximum flood scenario. As for the potential for human
wildlife conflict, two clusters emerge (\Cref{Metrics}c). The first cluster lies
at the inflow of the Okavango Delta between source areas 4 and 5 and is most
pronounced during minimum flood (\Cref{Difference}c). Another, albeit visually
less distinct, cluster covers the area at the distal end of the OD, stretching
from lake Ngami to Maun. This area appears particularly relevant at maximum
flood (\Cref{Difference}c). Our analysis of inter-patch connectivity further
demonstrates notable differences in dispersal prospects and dispersal durations
depending on the extent of the flood (\Cref{Metrics}d and \Cref{IPCTable}).
While \input{99_NumberReachingOthersMinFlood.tex} simulated dispersers reach
another source area during minimum extent, only
\input{99_NumberReachingOthersMaxFlood.tex} do so during maximum extent, thus
indicating an overall decrease in dispersal success of
\input{99_NumberReachingOthersPercentage.tex}\% during maximum flood.
Concomitantly, the average minimum dispersal durations increases by
\input{99_StepsToReachingOthersPercentage.tex}\%, i.e. from
\input{99_StepsToReachingOthersMinFlood} steps to
\input{99_StepsToReachingOthersMaxFlood} steps during maximum flood. These
differences are particularly pronounced for individuals dispersing into source
area 6 on Chief's Island. While the area is reached by
\input{99_NumberReachingArea6MinFlood.tex} simulated individuals during minimum
flood, only \input{99_NumberReachingArea6MaxFlood.tex}, i.e.
\input{99_NumberReachingArea6Percentage.tex}\% less, arrive there during maximum
flood. Furthermore, the dispersal duration into source area six from any other
source area increases by \input{99_StepsToReachingArea6Percentage}\% from
\input{99_StepsToReachingArea6MinFlood.tex} steps to
\input{99_StepsToReachingArea6MaxFlood.tex} steps. In few occasions,
connectivity between some areas increased during maximum flooding, for instance.
Temporary emigration increased from \input{99_TemporaryEmigrationMinFlood.tex}
to \input{99_TemporaryEmigrationMaxFlood.tex} trajectories (i.e.
by \input{99_TemporaryEmigrationPercentage.tex}\%). Permanent emigration increased
from \input{99_PermanentEmigrationMinFlood.tex} to
\input{99_PermanentEmigrationMaxFlood.tex} trajectories (i.e.
by \input{99_PermanentEmigrationPercentage.tex}\%).

\begin{figure}
 \begin{center}
  \includegraphics[width = \textwidth]{99_ImmigrationEmigration.png}
  \caption{Number of individuals emigrating from, or immigrating into a specific
  source area (focal area). Colors indicate into which other areas emigrants
  moved or from which other areas immigrants originate. For instance, the most
  left plot in the upper panel shows the number of individuals moving from
  source area 1 into the six other source areas during minimum and maximum
  flood, respectively.}
  \label{EmigrationImmigration}
 \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Emigration.png}
  \caption{Absolute number of simulated trajectories running into each of the
  designated emigration zones (purple) during minimum and maximum flood.
  Subfigure (a) depicts the total number of emigrating trajectories, including
  temporary emigrants that eventually returned into the OD's vicinity. Subfigure
  (b) only depicts permanent emigration events.}
  \label{Emigration}
  \end{center}
\end{figure}

@@

\section{Discussion}
\subsection{Brief Summary}
In this study, we used a previously parameterized and validated movement model
to simulate dispersal trajectories of AWDs across the OD under two extreme
environmental scenarios: one representing minimal flooding and one representing
maximum flooding. This approach allowed us to investigate connectivity patterns
that emerge under extreme environmental conditions, similar to those projected
under climate change. Predictions of flood conditions across the OD under
climate change remain ambiguous, yet it is generally agreed that climate change
will amplify climatic variability and result in either exacerbated or attenuated
flood events. Our two reference scenarios served to approximate these
conditions. By providing a comprehensive set of connectivity maps for both
scenarios, we highlighted how dispersal routes and prospects of the endangered
AWD changed depending on flood conditions. Our simulations revealed that the
propensity to move between the eastern and western sections of the OD decreased
significantly during maximum flood. This effect likely resulted from the
combined influence of floodwaters and anthropogenic pressures, which together
formed a dispersal barrier that limited connectivity. When flooding was at a
minimum, on the other hand, the retracted floodwaters revealed vital dispersal
habitats that facilitated movement between the western and eastern regions of
the OD. Anecdotal evidence supports the notion that vital dispersal habitats are
only available during periods of low flood, for the only dispersal coalition
recorded to successfully move between the eastern and western delta using GPS
data was observed at a time of minimal flooding  \citep{Cozzi.2020}. The lack of
dispersal habitat during maximum flood also resulted in an almost complete
isolation of Chief's Island, the OD's central peninsular (source area 6 in
\Cref{StudyArea}b). The peninsular itself remains remains dry in both scenarios,
yet becomes entirely surrounded by water during times of maximum flood. This
limits pathways to emigrate or immigrate and results in low dispersal prospects
for individuals moving towards this area. Despite the general reduction of
connectivity and increased dispersal durations during maximum flood, the number
of dispersers moving between some of the source areas increased during maximum
flood. This was, for instance, the case for movements between sour areas 1 and 4
and likely resulted from individuals that were deflected by the expanded flood
and redirected into more accessible habitats. The deflection of individuals by
the flood also had direct implications for the spatial arrangement of dispersal
corridors as movement routes that traverse the central parts of the OD
disappeared and more individuals were funneled through a corridor running
along the southern fringes of the OD.

\subsection{Validating Predictions}
Although the movement model underlying our simulations was validated using
independent dispersal data (see \citealp{Hofmann.2023}), assessing the
reliability of our predictions for the two extreme flooding scenarios remains
challenging. Firstly, because collecting data of dispersing individiauls is
difficult \textit{per se} \citep{Fattebert.2015, Cozzi.2020} and, secondly,
because flood-conditions akin to those studied here only occur rarely
\citep{Wolski.2017}. Dispersers typically only make up for a small proportion of
the entire population and predicting the timing of dispersal is often
non-trivial. Coupled with the limited battery-lifetime of most conventional GPS
collars, it is logistically unfeasible to collected large amounts of GPS data
during dispersal. However, as an alternative to GPS data, genetic or
observational data could also yield insight into functional connectivity.
Genetic data is often viewed as ultimate measure of functional connectivity
\citep{Baguette.2013} and thus often serves as validation of connectivity maps
(e.g. \citealp{Cushman.2010} or \citealp{Spear.2010}). Genetic analysis across
southern Africa revealed moderate levels of dispersal and identified a
genetically particularly diverse population cluster located near northern
Botswana \citep{Tensen.2022}. While such analyses provide valuable insights into
long-term dispersal patterns, they are unlikely to yield insights on short-term
connectivity patterns such as those studied here. Observational data, on the
other hand, readily delivers information on seasonally changing dispersal
patterns. Such data may not only be collected by trained field assistants, but
could also be obtained through photographic evidence from tourists in a
citicen's science approach. AWDs, as well as most other large carnivores, are
individudally identifiable, either by their coat pattern or other unique
features. Given a set of georeferenced images, individuals can thus be traced
through space and time. \cite{Cozzi.2013} provide first evidence on the
usefulness of this approach. As of today, however, the moderate amount of data
collected and spatially biased sampling effort prohibits using such data for
validation purposes. Finally, our representation of the flood was, by design,
focused on rare extreme events. Since the beginning of our dataset.
Consequently, the amount of data suitable for validation is limited.

>>>>

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Dominik M. Behr\textsuperscript{1,2} \orcid{0000-0001-7378-8538} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \orcid{0000-0001-7477-2642} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940}
}

@@

\textsuperscript{2} Botswana Predator Conservation Program, Private Bag 13,
Maun, Botswana.

@@

\textbf{Running Title:} Seasonal Flooding of the Okavango Delta and its
Consequences for African Wild Dog Dispersal and Connectivity

\textbf{Running Title:} Seasonal Flooding of the Okavango Delta and its
Consequences for African Wild Dog Connectivity under Climate Change

@@

\textbf{Keywords:} movement, connectivity, climate change, seasonality,
dispersal, conservation, individual-based

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Climate change is expected to profoundly impact the life history of wild-living
animal populations. While the impacts of climate change on the demographics of
local subpopulations have been studied repeatedly, little is known about the
consequences of environmental change on dispersal and connectivity.

We capitalize on a ``natural experimental setup'', the flood-pulse driven change
in surface-water across the Okavango Delta in northern Botswana, to investigate
the impact of changing environmental conditions on dispersal patterns and
connectivity of the endangered African wild dog (\textit{Lycaon pictus}). For
this, we simulate dispersal trajectories across the Okavango Delta under two
extreme environmental scenarios; one assuming a maximum flood, one assuming a
minimum flood. These scenarios serve to approximate environmental conditions
under amplified climatic variability, as it is expected under continued climate
change.

During maximum flood, the Okavango Delta poses an important dispersal barrier
that reduces dispersal prospects between distinct areas and increases dispersal
durations. At maximum flood, dispersal into the central habitats of the Okavango
Delta is reduced by $\input{99_NumberReachingArea6Percentage.tex}$\% and
dispersal durations increased by
$\input{99_StepsToReachingArea6Percentage.tex}$\%. Across the entire study area,
the differences are less dramatic, with
$\input{99_NumberReachingOthersPercentage.tex}$\% lower dispersal success and
$\input{99_StepsToReachingOthersPercentage.tex}$\% longer dispersal durations
during maximum flood. We also found that a maximum flood shifted dispersal
bottlenecks and areas with a high potential for human-wildlife conflict closer
towards densely populated villages. We provide a rich suite of maps to visualize
the differences between minimum and maximum flood.

Whilst the exact impacts of climate change on the flooding regime of the
Okavango Delta remain unknown, our results suggest that connectivity will vastly
differ depending on future flood conditions. Considering such differences will
be key to designing future proof conservation strategies, particularly in light
of ongoing climate change. Our results also have direct applicability for the
conservation of the endangered African wild dog as we highlight critical
dispersal corridors and human-wildlife conflict zones in two distinct scenarios.

@@

\subsection{What We Did}
Here, we utilize a previously parameterized and validated dispersal model as
IBMM to simulate dispersal and assess dispersal success and connectivity
patterns for African wild dogs under two extreme scenarios: one assuming maximum
flooding of the Okavango delta and one assuming minimum flooding of the Okavango
delta. The IBMM was trained using GPS data collected during dispersal and
frequently updated environmental data, thus providing a high degree of realism
\citep{Hofmann.2021}. Given that dispersers avoid crossing through water (albeit
we do occasionally observe it in the field), we anticipated that dispersal
prospects and connectivity during maximum flood would be low. Moreover,  when
the flood extent of the OD is at its maximum, the water extends almost into the
densely populated village of Maun. Since both the flood and humans are avoided
by dispersing wild dogs, we anticipated that a fully flooded Delta would result
in a total halt of movement between the Western and Eastern side of the OD.

@@

\subsection{Spatial Habitat Layers}
We represented the physical landscape through which dispersers could move by a
set of spatially referenced habitat layers, each with a resolution of 250 m. The
set of layers included water-cover, distance-to-water, tree-cover,
shrub/grassland-cover, and a human influence layer depicting anthropogenic
influences through villages, roads, and agriculture. A detailed description of
the different habitat layers is provided in \citep{Hofmann.2021}. Importantly,
the water-cover and derived distance-to water layers were generated using MODIS
Terra MCD43A4 satellite imagery that was classified into binary water-cover maps
using a ``floodmapping'' algorithm developed by \citep{Wolski.2017}. This
allowed us to generate almost weekly updated ``floodmaps'', thus providing
detailed information about the flood-extent at any given point in time. In
total, we generated 700 floodmaps between the years 2000 and 2019, based on
which we generated a minimal and maximum flood scenario. To create the minimum
(maximum) flood scenario, we averaged the 50 floodmaps with the smallest
(highest) flood extent and generated a binary layer by masking all pixels that
were inundated in less than 50\% of the maps. The resulting maps are presented
in \Cref{FloodExtent}. Ultimately, we combined the habitat layers into two
stacks, one representing the minimum flood scenario, one representing the
maximum flood scenario. To mitigate edge effects during the dispersal
simulation, we followed \cite{Koen.2010} and expanded the spatial extent of the
stacked layers by 20\% and randomized habitat values within the so created
buffer zone (red rectangle in \Cref{StudyArea} a).

@@

\subsection{Dispersal Simulation}
We used a previously parameterized and validated dispersal model to simulate
dispersal of AWDs. The dispersal model was trained using GPS data of 16 wild dog
coalitions dispersing across northern Botswana \citep{Hofmann.2023} which was
fed into an integrated step-selection function (iSSF, \citealp{Avgar.2016}). In
iSSFs, consecutive GPS locations are converted into steps (the straight-line
traveled between two GPS recordings \citep{Turchin.1998a}) and compared to a set
of \textit{random} steps in a conditional logistic regression framework
\citep{Fortin.2005, Thurfjell.2014, Muff.2020, Fieberg.2021}. The model
presented in \cite{Hofmann.2023} comprised of a movement kernel, describing how
dispersers move across the landscape in the absence of habitat selection, a
habitat kernel, indicating preferred or avoided habitat features, and
interactions among the two, i.e. how movement behavior changes depending on
habitat conditions. According to this model, the main characteristics of AWD
dispersal movements are avoidance of water, avoidance of areas influenced by
humans, and a preference for directional and fast movements. The model
parameters are provided in Appendix SX and explained in \citealp{Hofmann.2023}.

Originating from each of the six source areas, we simulated 2'000 individuals
dispersing for a total of 2'000 steps. 1'000 individuals were simulated assuming
a minimum flood, the remaining 1'000 assuming a maximum flood. This resulted in
the simulation of a total of 12'000 individuals. The simulation procedure was
based on the algorithm described in \Cref{Hofmann.2023} and works as follows. A
random location within the source area is defined as starting point. Originating
from the starting point, a set of 25 random steps is generated by sampling step
lengths from a gamma distribution fitted to observed steps (shape = 0.37, scale
= 6'316) and turning angles from a uniform distribution ($-\pi, +\pi$). Along
each random step the underlying spatial covariates are extracted, and relevant
movement metrics are computed. \(\beta-\)estimates from the fitted model are
used to predict the probability of each step for being chosen, given the steps
associated covariates. Among the 25 proposed steps, one is chosen at random
based on assigned probabilities. The location of the animal is updated, and the
procedure is repeated until the desired number of steps is realized. Here, we
simulated each individual for 2'000 steps, corresponding to a dispersal duration
of 400 days and the longest dispersal duration recorded in this study area
\citep{Cozzi.2020, Hofmann.2021}. The simulated trajectories can be understood
as correlated random walks.

@@

\section{Results}
Figures depicting the derived connectivity and human-wildlife conflict maps are
provided in \Cref{Metrics}. Difference maps to visualize the differences between
minimum and maximum flood are given in \Cref{Difference}. Source-specific maps
are presented in the Appendix. As the heatmaps in \Cref{Metrics}a reveal, the OD
acts as major dispersal barrier during maximum flood yet reveals vital
dispersal habitat during minimum flood. Differences between maximum and minimum
flood (\Cref{Difference}a) are particularly pronounced for the region between
source areas 1 and 2, where few dispersers occur during times of maximum flood.
In fact, because the floodwaters of the OD reach almost into Maun, the OD
creates a line of separation between its eastern and western sections. The
separation is further amplified as the city of Maun is avoided by dispersers in
both scenarios. Similar patterns are observed on the betweenness maps
(\Cref{Metrics}b), where several pinch-points and bottlenecks linking source area
6 to the surrounding source areas exist during minimum flood. During maximum
flood, however, these links vanish and instead a single corridor at the
south-eastern tip of the OD emerges (\Cref{Difference}b). Despite its apparent
importance in linking the eastern and western sections of the delta, it is
evident from (\Cref{Metrics}a) that this corridor is only rarely used,
especially during the maximum flood scenario. As for the potential for human
wildlife conflict, two clusters emerge (\Cref{Metrics}c). The first cluster lies
at the inflow of the Okavango Delta between source areas 4 and 5 and is most
pronounced during minimum flood (\Cref{Difference}c). Another, albeit visually
less distinct, cluster covers the area at the distal end of the OD, stretching
from lake Ngami to Maun. This area appears particularly relevant at maximum
flood (\Cref{Difference}c). Our analysis of inter-patch connectivity further
demonstrates notable differences in dispersal prospects and dispersal durations
depending on the extent of the flood (\Cref{Metrics}d and \Cref{IPCTable}).
While $\input{99_NumberReachingOthersMinFlood.tex}$ simulated dispersers reach
another source area during minimum extent, only
$\input{99_NumberReachingOthersMaxFlood.tex}$ do so during maximum extent, thus
indicating an overall decrease in dispersal success of
$\input{99_NumberReachingOthersPercentage.tex}$\% during maximum flood.
Concomitantly, the average minimum dispersal durations increases by
$\input{99_StepsToReachingOthersPercentage.tex}$\%, i.e. from
$\input{99_NumberReachingOthersMinFlood}$ steps to
$\input{99_NumberReachingOthersMaxFlood}$ steps during maximum flood. These
differences are particularly pronounced for individuals dispersing into source
area 6. While the area is reached by
$\input{99_NumberReachingArea6MinFlood.tex}$ simulated individuals during
minimum flood, only $\input{99_NumberReachingArea6MaxFlood.tex}$, i.e.
$\input{99_NumberReachingArea6Percentage.tex}$\% less, arrive there during
maximum flood. Furthermore, the dispersal duration into source area six from any
other source area increases by $\input{99_StepsToReachingArea6Percentage}$\%
from $\input{99_StepsToReachingArea6MinFlood.tex}$ steps to
$\input{99_StepsToReachingArea6MaxFlood.tex}$ steps.

@@

\section{Discussion}
\subsection{Brief Summary}
According to our simulations, the propensity to move between the eastern and
western part of the delta is much lower during maximum extent. This is mainly
due to the flood-waters and the city of Maun acting as dispersal barriers.
During maximum extent, the floodwaters of the delta close a gap between the
delta and Maun that otherwise would serve as dispersal corridor. Anecdotal
evidence supports this hypothesis, for the only dispersing individuals recorded
to move from the eastern to the western part of the delta moved at times of low
flood. In line with this, it appears that a large flood extent pushes dispersing
individuals to move closer to human inhabited areas such as the village of Maun.

Predicting how climate change will impact the dispersal ability of AWDs is
challenging for multiple reasons. First of all, predicting the flooding patterns
of the OD under climate change is non-trivial due to complex feedback-loops
between surface-temperature, soil conditions, precipitation patterns, and the
associated changes in vegetation.

Second, the delta is not only prone to changes
in environmental conditions, but also to changes in anthropogenic use of the
inflowing water. Finally, it is unclear how AWDs, in fact, how any
species, will cope with environmental change due to global warming. Even though
some studies predict that AWD populations are likely to decline under increasing
temperatures, these studies fail to account for the behavioral plasticity of
their focal species. AWDs respiratory system, for instance, has evolved as a
perfect adaptation to high temperatures and AWDs may, in fact, profit from a
comparative advantage (cite an economist) over their competitors and prey under
rising temperatures. Although the theory of comparative advantages is a
fundamental concept in economics, it has yet to find its way into ecological
studies.

\cite{Murray-Hudson.2006} predicted that increased temperatures, additional human
abstractions, and reduced river flows might lead to a ``Delta dying'' and that
the impact of climate will be much more pronounced than the impact of
anthropogenic water use. However, the paper also found that predicted conditions
vary from much dryer to much wetter depending on the employed climate model.

While precipitation in the OD's catchment areas are expected to increase, it
remains unclear wheather the effect will be offset by increased
evapotranspiration. \cite{Wolski.2018} predict changes that can range from much
``wetter'' to much  ``drier'' conditions depending on the employed climate
model. In conclusion, there is too much uncertainty to make reliable
predictions, meaning that conservation practicioners will need to retain
flexibility in their planning.

Climate models consistently predict an above global average increase in
temperatures for Southern Africa.

There is large uncertainty regarding the future anthropogenic use and
abstractions from the Okavango river \citep{Hughes.2011}.

It has been suggested that climate change will intensivy human wildlife conflict
due to increased competition for scarce resources \cite{Abrahms.2021}. Instead
of \textit{increased} human wildlife conflict, our results suggest that human
wildlife conflict will shift. Given that these areas have had little exposure to
such conflicts in the past, inexperience and lack of knowledge may explain the
resulting apparent increases in conflict situations.

Besides the directional change in climate, it is agreed that there will be an
increased frequence in extreme climate events.

Along a similar vein, the shrinking of the sea-ice has resulted in a higher
incidence of problem polar bears in Canada over the past decades
\citep{Towns.2009}. Similarly, it has been found that the annual range shifts of
polar bears, in result to seasonally melting sea-ice, results in increased
frequency of HWC.

Step-selection functions (or path selection functions) provide better estimates
of connectivity than other resource selection approaches \citep{Zeller.2016}.

Information on habitat selection or connectivity are suitable for predicting
areas with an elevated potential for human wildlife conflict
\citep{Buchholtz.2020}.

Although local rainfalls in Botswana are expected to increase in terms of
intensity, simulations show that the length of the rainy season will decline,
more than offsetting the incline in precipitation \cite{Akinyemi.2019}.

Although dispersal is generally believed to be beneficial for the long term
viability of species, it can also lead to the net-loss of individuals in areas
where edge effects and anthropogenic mortality are high \citep{Leigh.2012}.

We studied a population of African wild dogs that resides in a natural
environment with little human influence. This is only representative for a small
share of the extant wild dog populations, as most individuals reside in areas
that are prone to substantial edge effects. For these populations, the benefit
of dispersal is disputed, as dispersal under high anthropogenic mortality may
lead to a net-loss of individuals \citep{Leigh.2012}.

Dispersal is an important, if not the most important, driver of landscape
connectivity and therefore of major interest to conservation authorities. It has
also been demonstrated that dispersal may lead to population declines in areas
where anthropogenic mortality is high and dispersal prospects low
\citep{Leigh.2012}.

Efforts aiming at restoring or improving landscape connectivity should consider
and take into account changes in landscape characteristics due to climate
change.

Maybe cite the \cite{Gusset.2009} paper.

We assessed the implications of environmental change on the dispersal prospects,
yet we did not consider how changing conditions alter dispersal propensity.

We studied the indirect (through the landscape), but not the direct (propensity
to disperse) conesquences of climate change on dispersal.

According to our simulations, dispersers are able to cover larger distances
during periods of low flood. This finding is little surprising, considering that
inundated areas act as dispersal barriers and force dispersers to detour and
circumvent water-covered areas. However, it still leads to an interesting
hypothesis. Previous studies have shown that the euclidean dispersal distance of
female coalitions is larger than that of male coalitions. This has led to ...
However, demographic analyses have also revealed that female offspring tend to
emigrate from their pack at younger ages and earlier in the year, when
floodwaters are still at a relatively low level (Behr). It is thus conceivable
that the sex-differences in reported dispersal distances is mainly a consequence
of environmental conditions during dispersal, rather than owed to physiological
differences between sexes.

Unsurprisingly, \cite{Michalski.2006} show that increased proximity between
wildlife and humans significantly increases the likelihood of human-wildlife
conflict.

\subsection{Effect of Climate Change on the Okavango Delta}
\cite{Akinyemi.2019} predicts fewer rain days in Botswana (may want to look at
predictions for Angola though)

While a better understanding of the ODs flooding regime under global change is
of major interest, the specific impacts of climate change on the OD are not
fully understood. This is owed to the complex interplay between climate,
anthropogenic water use, animal behavior, and topography of the region. Talk a
bit about anthropogenic water use (e.g. Namibia is building a hydropower station
in the Zambezi Region) Talk a bit about droughts, etc. Rising sea-level ->
Intrusion of salt water Simulations by \cite{Wolski.2006} reveal that upstream
abstractions have a smaller infuence on flooding patterns of the OD than climate
change. In combination, however, Simulated upstream abstractions and climate
change resulted in significant ``delta-dying'', i.e....

\subsection{Species Distribution}
Climate change and anthropogenic use will result in trophic ripple effects
\citep{Wolski.2006}. Climate change affects species distribution through habitat
suitability, which determines structural connectivity which ultimately
determines functional connectivity. How climate change will affect the
distribution of African wild dogs is not known. Given that the species has
occured across the entire Sub-Saharan continent suggests that this species is
highly adaptable and limited mainly by the presence of humans, yet less by the
habitat itself. In fact, wild dogs have been found on mount Kilimandjaro.

\subsection{Social Landscape}
The OD is an important driver of species distribution and it has been found that
an expanding flood limits available habitat, thus leading to more inter-specific
competition, particularly between AWDs and lions (Robynne).
Although the OD is arguably the main driver of seasonal change across the
studied ecosystem, there are several other factors that undergoe seasonal
change, including vegetation and the abundance and distribution of prey or
predators.An additional complication arises when species movement is not solely driven by
environmental conditions, but also affected by itra- and inter-specific factors.
For instance, ... has shown that dispersers...
Rendering such conditions alone is challenging, yet rendering the conditions
under changing environmental conditions is merely impossible.
As highlighted in the papers presented in the intro, climate change change
environmental conditions, and in return species distribution, which will, may
ultimately culminate in alterations of community composition
\citep{Thuiller.2006}. Previous studies have shown that intra-specific factors
are imortant determinants of dispersal movements \citep{Cozzi.2018} and it can
be expected that inter-specific factors are similarly important
\citep{Armansin.2019}. Through changing environmental conditions, climate change
will also impact the distribution of various species with far-reaching
consequences on inter- and intra-specific competition (Abrahms). Due to a lack
of data, we omitted any inclusion of social factors in our dispersal model,
albeit they are known to be important determinants of dispersal (Armansin).
Studying how climate change influences species distribution and their
interactions, and ultimately how this alters dispersal, will be challenging but
necessary to more realistically simulate dispersal.

\subsection{Seasonality in Dispersal Models}
While our analysis marks an important step into incorporating environmental
change into studies of connectivity, there are several critical additions that
should be considered in the future. We studied dispersal and connectivity under
two different environmental scenarios, yet our movement model assumed that
dispersers had identical habitat and movement preferences in both scenarios. In
reality, however, it can be expected that movement and habitat kernels of
dispersers differ depending on the season considered (examples).
Can incporporate seasonality by keeping environmental conditions constant, but
rendering changes in species habitat preferences. Or, keep species habitat
preferences constant and vary the environemnt. Or both. To address such
differences, researchers could model habitat and movement preferences using
season-dependent models, or, alternatively, by combining hidden markov models
with step-selection functions. (cite papers that fieberg sent)
